---
title: "Experiments with parameter mixing for Adrakey model"
author: "Adam Zeiliinger"
date: "6/12/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Results from examining mixing and estimation problems with Adrakey model parameters

Below are the results for three "experiments". First, I set each of the three parameters (alpha, beta, and epsilon) equal to the "true value", where the true value is the one used by Adrakey et al. to run their simulation in the Suppl. Material. Then I removed monitors on two of the three parameters.

The true values of the parameters are:  
  - alpha = 0.08  
  - beta = 7e-06  
  - epsilon = 5e-05  

I ran each experiment with the following MCMC conditions:  
  - iterations = 100K   
  - burnin = 1000  
  - thin = 2  
  - nchains = 2  
  - RNG seed = 1 (seed set differently for different MCMC chains, but only chain 1 analyzed below)


```{r, echo = FALSE, include = FALSE}

my.packages <- c("dplyr", "tidyr", "ggplot2", "coda", "data.table")
lapply(my.packages, require, character.only = TRUE)


```



## Results of experiment monitoring alpha

```{r, echo=TRUE}

samples <- readRDS("output/raw_mcmc_samples_alpha_test.rds")
samplesAlpha <- samples[[1]]


## Getting just the model parameters
paramcols <- which(attr(samplesAlpha, "dimnames")[[2]] == "alpha")
paramsAlpha <- samplesAlpha[,paramcols]
## Make paramList a list of MCMC samples of just the three parameters
paramAlphaList <- list(paramsAlpha, samples[[2]][,paramcols])

plot(paramsAlpha)

## Gelman-Rubin diagnostic
gelman.diag(paramAlphaList)

## Effective sample size
round(cbind(
  niterations = length(paramsAlpha),
  ESS    = effectiveSize(paramsAlpha)
))

## Summary of posterior distributions
res <- round(cbind(
  `cilower` = apply(samplesAlpha, 2, function(x) quantile(x, 0.025)),
  `mean`    = apply(samplesAlpha, 2, mean),
  `median`  = apply(samplesAlpha, 2, median),
  `ciupper` = apply(samplesAlpha, 2, function(x) quantile(x, 0.975))
), 8)

tail(res)


```



## Results of experiments monitoring beta

```{r, echo = TRUE}

samples <- readRDS("output/raw_mcmc_samples_beta_test.rds")
samplesBeta <- samples[[1]]


## Getting just the model parameters
paramcols <- which(attr(samplesBeta, "dimnames")[[2]] == "beta")
paramsBeta <- samplesBeta[,paramcols]
## Make paramList a list of MCMC samples of just the three parameters
paramBetaList <- list(paramsBeta, samples[[2]][,paramcols])

plot(paramsBeta)


## Burnin apparently not long enough for beta
## Try plotting again after removing an additional 1000 iterations
paramsBeta <- paramsBeta[1001:length(paramsBeta)]
plot(as.mcmc(paramsBeta))


## Gelman-Rubin diagnostic
gelman.diag(paramBetaList)

## Effective sample size
round(cbind(
  niterations = length(paramsBeta),
  ESS    = effectiveSize(paramsBeta)
))

## Summary of posterior distributions
res <- round(cbind(
  `cilower` = apply(samplesBeta, 2, function(x) quantile(x, 0.025)),
  `mean`    = apply(samplesBeta, 2, mean),
  `median`  = apply(samplesBeta, 2, median),
  `ciupper` = apply(samplesBeta, 2, function(x) quantile(x, 0.975))
), 8)

tail(res)


```



## Results of experiments monitoring epsilon

```{r, echo = TRUE}

samples <- readRDS("output/raw_mcmc_samples_epsilon_test.rds")
samplesEpsilon <- samples[[1]]


## Getting just the model parameters
paramcols <- which(attr(samplesEpsilon, "dimnames")[[2]] == "epsilon")
paramsEpsilon <- samplesEpsilon[,paramcols]
## Make paramList a list of MCMC samples of just the three parameters
paramEpsilonList <- list(paramsEpsilon, samples[[2]][,paramcols])

plot(paramsEpsilon)

## Gelman-Rubin diagnostic
gelman.diag(paramEpsilonList)

## Effective sample size
round(cbind(
  niterations = length(paramsEpsilon),
  ESS    = effectiveSize(paramsEpsilon)
))

## Summary of posterior distributions
res <- round(cbind(
  `cilower` = apply(samplesEpsilon, 2, function(x) quantile(x, 0.025)),
  `mean`    = apply(samplesEpsilon, 2, mean),
  `median`  = apply(samplesEpsilon, 2, median),
  `ciupper` = apply(samplesEpsilon, 2, function(x) quantile(x, 0.975))
), 8)

tail(res)


```


## Results from mixing experiments employing multiple configurations

For these runs, I set up two MCMC configurations:  
  - MCMCconf: where samplers for all three parameters (alpha, beta, epsilon) were removed, so only infection times were sampled  
  - MCMCconf2: where samplers were removed for 2 out of 3 parameters; here infection times and one parameter (either beta or epsilon) were sampled

I ran the MCMCs with these configurations in serial. I specified 5,000 iterations for the first run (for MCMCconf) and 100,000 iterations for the second run (for MCMCconf2). Otherwise, the specifications for the MCMCs were identical. I did not set a RNG seed prior to either MCMC run.
  

```{r, eval = FALSE, include = TRUE}

MCMCconf <- configureMCMC(Rmodel, nodes = c("alpha", "beta", "epsilon"))
MCMCconf$addSampler(type = 'sampler_infection_quiet', 
                    target = 'Inf_indices',
                    control = list(
                      time_intervals = Y,
                      Inf_times_node = 'Inf_times',
                      Inf_indices_node = 'Inf_indices',
                      numInfections_node = 'numInfections',
                      latent_period_node = 'latent_period',
                      Tmax_node = 'Tmax'
                    ))
MCMCconf$addMonitors('Inf_times')
MCMCconf$removeSamplers(c('alpha', 'beta', 'epsilon'))
MCMC <- buildMCMC(MCMCconf)
MCMCconf$printSamplers()



MCMCconf2 <- configureMCMC(Rmodel, nodes = c("alpha", "beta", "epsilon"))
MCMCconf2$addSampler(type = 'sampler_infection_quiet', 
                    target = 'Inf_indices',
                    control = list(
                      time_intervals = Y,
                      Inf_times_node = 'Inf_times',
                      Inf_indices_node = 'Inf_indices',
                      numInfections_node = 'numInfections',
                      latent_period_node = 'latent_period',
                      Tmax_node = 'Tmax'
                    ))
MCMCconf2$addMonitors('Inf_times')
MCMCconf2$removeSamplers(c('alpha', 'beta'))
MCMC2 <- buildMCMC(MCMCconf2)
MCMCconf2$printSamplers()

## Compile MCMC for running in C++
Cmcmc <- compileNimble(MCMC, MCMC2, project = Rmodel, resetFunctions = TRUE)


####################################################################################################
#### Multiple MCMC run on Adrakey simulated data

nburnin <- 1
thin <- 2
nchains <- 1

samples <- runMCMC(Cmcmc$MCMC, niter = 5000, nburnin = nburnin, thin = thin, 
                   nchains = nchains, samplesAsCodaMCMC = TRUE)
samples2 <- runMCMC(Cmcmc$MCMC2, niter = 100000, nburnin = nburnin, thin = thin, 
                    nchains = nchains, samplesAsCodaMCMC = TRUE)



```


## Results of the multi-configurations experiments:
### Epsilon sampled; examining only the posterior of the 2nd MCMC run (samples2), where epsilon was sampled along with infection times 


```{r, echo = TRUE}

samples <- readRDS("output/raw_mcmc_samples_multi-conf_epsilon_test_in_serial.rds")
samplesMCepsilon <- samples[[2]]


## Getting just the model parameters
paramcols <- which(attr(samplesMCepsilon, "dimnames")[[2]] == "epsilon")
paramsEpsilon <- samplesMCepsilon[,paramcols]

plot(paramsEpsilon)


## Apply a burnin and plot again
paramsEpsilon <- paramsEpsilon[5001:length(paramsEpsilon)]
plot(as.mcmc(paramsEpsilon))


## Effective sample size
round(cbind(
  niterations = length(paramsEpsilon),
  ESS    = effectiveSize(paramsEpsilon)
))

## Summary of posterior distributions
res <- round(cbind(
  `cilower` = apply(samplesMCepsilon, 2, function(x) quantile(x, 0.025)),
  `mean`    = apply(samplesMCepsilon, 2, mean),
  `median`  = apply(samplesMCepsilon, 2, median),
  `ciupper` = apply(samplesMCepsilon, 2, function(x) quantile(x, 0.975))
), 8)

tail(res)



```



## Results of the multi-configurations experiments:
### Beta sampled; examining only the posterior of the 2nd MCMC run (samples2), where beta was sampled along with infection times 


```{r, echo = TRUE}

samples <- readRDS("output/raw_mcmc_samples_multi-conf_beta_test_in_serial.rds")
samplesMCbeta <- samples[[2]]


## Getting just the model parameters
paramcols <- which(attr(samplesMCbeta, "dimnames")[[2]] == "beta")
paramsBeta <- samplesMCbeta[,paramcols]

plot(paramsBeta)


## Apply an additional burnin and plot again
paramsBeta <- paramsBeta[5001:length(paramsBeta)]
plot(as.mcmc(paramsBeta))


## Effective sample size
round(cbind(
  niterations = length(paramsBeta),
  ESS    = effectiveSize(paramsBeta)
))

## Summary of posterior distributions
res <- round(cbind(
  `cilower` = apply(samplesMCbeta, 2, function(x) quantile(x, 0.025)),
  `mean`    = apply(samplesMCbeta, 2, mean),
  `median`  = apply(samplesMCbeta, 2, median),
  `ciupper` = apply(samplesMCbeta, 2, function(x) quantile(x, 0.975))
), 8)

tail(res)



```


## Summary

In summary, the results of the multi-configuration experiments were equivalent to the first set of experiments. The Effective Sample Size and posterior distributions were unaffected after initializing the sampling of infection times with the first MCMC run.


## Results of experiments from 2019-09-12

The following experiments were run with the following MCMC conditions:  
  - iterations = 20K   
  - burnin = 1000  
  - thin = 1  
  - nchains = 1  

#### Results from experiment with Tmax = 460 

```{r, echo = TRUE}

samples <- readRDS("output/raw_mcmc_samples_test_2019-09-12.rds")
samplesTmax <- samples


## Getting just the model parameters
paramcols <- which(attr(samples, "dimnames")[[2]] == "alpha" | attr(samples, "dimnames")[[2]] == "beta" | attr(samples, "dimnames")[[2]] == "epsilon")
## Need to remove an additional burnin
paramsTmax <- as.mcmc(samplesTmax[,paramcols])

plot(paramsTmax)


## Effective sample size
round(cbind(
  niterations = nrow(paramsTmax),
  ESS    = effectiveSize(paramsTmax)
))

## Summary of posterior distributions
res <- round(cbind(
  `cilower` = apply(samplesTmax, 2, function(x) quantile(x, 0.025)),
  `mean`    = apply(samplesTmax, 2, mean),
  `median`  = apply(samplesTmax, 2, median),
  `ciupper` = apply(samplesTmax, 2, function(x) quantile(x, 0.975))
), 8)

tail(res)



```


## Summary from experiment setting Tmax = 460

Burn-in was not nearly long enough and led to lower mean estimates. The 95% CI are very large. Need to run it with more iterations.


#### Re-run experiment with 40K iterations and all other conditions the same

```{r}




```

