---
title: "Seasonal Pierce's Disease model"
author: "Adam Zeiliinger"
date: "11/6/2019"
output: html_document
---

```{r setup, include=FALSE}
my.packages <- c("ggplot2", "raster", "dplyr", "tidyr", "data.table")
lapply(my.packages, require, character.only = TRUE)

#### Source additional functions
## Vector density function
#source("R_functions/seasonal_vector_density_functions.R")
## Dispersal kernel functions, also includes makeCoordinates() function
source("R_functions/dispersal_kernel_functions.R")
source("R_functions/factor2numeric.R")
```

## Host compartments

The model consists of a spatially explicit, discrete-time SEIDS host compartments. Susceptible hosts become infected and move to the Exposed (or latently infected) compartment according to a force of infection $\lambda_{i}(t)$. After a latent period, host i moves into the I, or Infectious, compartment and can now contribute to transmitting the pathogen to other hosts (through vectors). However, hosts in the Infectious compartment are considered asymptomatic. Then after some post-latent incubation period, host i moves into the D, or Diseased, compartment where it remains infectious but now also symptomatic. This framework is based on Parry et al. (2014). In additon, hosts in any infected compartment (E, I, D) can recover and become Susceptible again.

\ 

### Adrakey model for force of infection

At its core, we are modeling the instantaneous force of infection of susceptible plant $i$ in the interval between time $t$ to $t + dt$, $\lambda_{i}(t)$, as a function of secondary or vine-to-vine spread rate $\beta$, dispersal scale parameter $\alpha$, and external or primary rate of infection $\epsilon$:

$$
\begin{align}
\lambda_i(t) &= \beta \sum_{j \in I(t)} K(d_{i,j},\alpha) + \epsilon \\
\ \\
\text{Dispersal kernel function is a normalized negative exponential} \\
\ K(d,\alpha) &= \frac{1}{2 \pi d \alpha} \cdot \exp(-\frac{d}{\alpha}) \\
\end{align}
$$

### Transitions among compartments are stochastic
#### Sellke threshold construction

Assume that each susceptible plant $i$ has a threshold of resistance $Q_{i}$ which are exponentially distributed and independent of each other: $Q_{i}\sim\exp(1)$. The cumulative pressure on plant $i$ by time $t$ is given by: 
$$A_i(t)=\int_{0}^{t}\lambda_{i}(u)du$$
Then plant $i$ becomes latently infected at time $t$ when $Q_{i}=A_{i}(t)$. In this way, we relate the force of infection upon a plant to its infection status in a stochastic manner (Adrakey et al. 2017; Sellke 1983).

#### Infectious and Diseased compartment transitions
The transition from Exposed to Infectious is determined by a plant-specific latent period which is assumed to take the form of a gamma distribution, $t_{i}^I|t_{i}^E \sim gamma(\lambda_{EI}, \nu_{EI})$. Likewise, the transition from Infectious to Diseased is treated the same: $t_{i}^D|t_{i}^I \sim gamma(\lambda_{ID}, \nu_{ID})$. The values for the shape parameters were both set to $\lambda_{EI}, \lambda_{ID} = 4$, while the values of the scale parameters were both set to $\nu_{EI}, \nu_{ID} = 1$. This produces a humped shaped distribution with a mean value of 4 weeks for each time period, meaning that, on average, plants move from Exposed to Diseased in 8 weeks.

```{r, fig.align='center', fig.cap='Gamma distribution for latent period and post-latent incubation period'}

## Durations of stays in Exposed compartment
Exposed_time <- rgamma(900, shape = 4, scale = 1)
hist(Exposed_time)

```


## Incorporating vector density and infectivity

Assume that the external infection rate $\epsilon$ and secondary infection rate $\beta$ are time-dependent functions of vector density and infectivity. This is inspired by Parry et al. (2014), who used a similar model to Adrakey et al., but deviates from their exact formulation.

#### External infection rate:

$$
\begin{align}
\text{Time-dependent external infection rate for plant i} \\
\epsilon_{i}(t) &= \eta \kappa \cdot \rho_{\epsilon}(t) \cdot K(D_{i}, \alpha) \\
\text{where } \rho_{\epsilon}(t) \text{ is a piecwise function as follows}: \\
\rho_{\epsilon}(t)= \left\{
\begin{array}{ll}
  \tilde{\rho}(t) & \tilde{\rho}(t)>0 \\
  \ 0 & otherwise \\
\end{array}
\right. \\
\text{with} \\
\tilde{\rho}(t) &= A \cdot \exp(-\lambda t) \cdot \sin(\phi \pi t) \\
\end{align}
$$

Here, $\eta$ is the vector inoculation rate, $\kappa$ is the proportion of external vectors that are infectious, $K(D_{i},\alpha)$ is the immigrant dispersal kernel function, and $\rho_{\epsilon}(t)$ is the average density of immigrating external vectors from border vegetation. The immigrant dispersal kernel function could take an equivalent form as the within-field dispersal function $K(d_{i,j},\alpha)$ or could take a different form. The important difference is that whereas $d_{i,j}$ is the distance from susceptible plant $i$ and infected plant $j$, the distance $D_{i}$ is the distance from plant $i$ to the nearest border of the field. This form assumes that immgrating vectors are more likely to land on a plant near the field border than further inside the field.

```{r}

## Example immigrant vector dispersal
alpha <- 10 # Dispersal parameter

#### Plant spatial coordinates
## Total number of plants (numPlants) = nrow(Coo) = nrc*nrc
nrc <- 50 # Number of rows/columns
Coo <- makeCoordinates(nrc)

## Calculating distance to nearest field border for each plant
borderCoo <- rep(c(1,nrc), each = 2)
borderD <- apply(Coo, 1, function(x) min(sqrt((x - borderCoo)^2))+1)
## Add 1 to the distance; plants on the border are then 1 space from the border and all other plants are shifted accordingly
## Dispersal distances from border
epsilonK <- chooseKernel(d = borderD, alpha = alpha, kernelFunc = "normalized exponential")
qplot(borderD,epsilonK,geom="path", xlab="Distance from border", ylab="K(D,alpha)")

```

\ 
The density of immigrating vectors (per plant), $\rho_{\epsilon}(t)$, takes the form of a damped sine wave. The idea is to replicate the generational peaks of BGSS in the border vegetation observed in North Coast data set. This is a pretty crude phenomenological approach and probably needs more detail. Nonetheless with the right parameter values we can get a good approximation of the dynamics within 1 year. The piecewise element is to ensure that the function doesn't produce negative densities; a value greater than zero could also be specified for the "baseline".

```{r}

vectorDensity <- function(A, lambda_osc, phi, base = 0, time){
  rho <- A*exp(-lambda_osc*time)*sin(phi*pi*time)
  ## set negative numbers to zero
  rho <- ifelse(rho < base, base, rho)
  return(rho)
}

timeVec <- seq(1,52,0.1)
rho_et <- vectorDensity(A = -10, lambda_osc = 0.05, phi = 0.09, base = 0, time = timeVec)
qplot(timeVec, rho_et, xlab = "Week", ylab = "Immigrating vector density", geom = "path")

```

\ 

#### Secondary infection rate


$$
\begin{align}
\text{Time-dependent secondary infection rate} \\
\beta(t) &= \eta \kappa_{\beta}(t) \cdot \rho_{\beta}(t) \\
\text{where} \\
\kappa_{\beta}(t) &= \frac{a(t-1)I(t-1)}{N} \\
\rho_{\beta}(t) &= (1 - \mu_{v}) \cdot (\rho_{\beta}(t-1) + \rho_{\epsilon}(t-1)) \\
a(t) &= \frac{bt^2}{c + dt + t^2}
\end{align}
$$

Here, $\eta$ is the same inoculation rate as above. But now $\kappa_{\beta}(t)$ is the time-dependent proportion of in-field vectors infectious, and is a function of the density of infected grapevines at the previous time step $\frac{I(t-1)}{N}$ and the time-dependent acquisition rate $a(t-1)$. The density of in-field vectors, $\rho_{\beta}(t)$, is a function of the density of in-field vectors at the previous time step, $\rho_{\beta}(t-1)$, the density of immigrating vectors at the previous time step, $\rho_{\epsilon}(t-1)$, and the loss rate of vectors, $\mu_{v}$, due to death and emmigration. This assumes that reproduction in field is negligible.

\ 

The acquisition rate $a(t-1)$ is a function of time according to a Holling Type IV equation; this approximately replicates the acquisition rates in Daugherty and Almeida 2019 Fig. 1B with the parameter values: a = 0.1, b = 72, c = -16. The acquisition rate should be a function of Xylella population growth in infected plants, which in turn should be a function of time of infection. We could construct a sub-model for acquisition rate that is based on the infection times, since the model tracks this. But it's not immediately clear how to do this.

\ 

```{r}
#### Time-dependent acquisition
holling4 <- function(a, b, c, time){
  (a*time^2)/(b + c*time + time^2)
}

a <- 0.1 # long-term y value
## x value at peak = -2b/c should coincide with Sept/Oct
b <- 1134
c <- -63
time <- seq(1,52,0.1)

at <- holling4(a, b, c, time)

qplot(time, at, geom="path", xlab="Week", ylab="Acquisition rate")

```

\ 

## Inter-annual processes

At the end of each growing season (i.e., winter), we assume that winter recovery, plant dormancy, plant roguing, and vector overwintering occur instantaeously. This is similar to the assumptions of Daugherty and Almeida (2019), Gruber and Daugherty (2013), and the impulse equations of Rozins and Day (2016). 

We assume that the probability of winter recovery for plant $i$, $P_{i}^R$, is a function of infection time (in day of year, DOY), specifically the time that it moved into the Exposed compartment during the growing season, according to the functional response from Feil et al. (2003):
$$P_{i}^R = (1 - e^{-0.045 \cdot DOY})^{135.9}$$

Recovery status of plant $i$ is then a random binomial variable with probability $P_{i}^R$. If plant i recovers, it then reverts back to Susceptible. If plant i does not recover it reverts back to the Exposed compartment (if it had moved into the Infectious or Diseased compartments within the year). Starting at the beginning of the next year, plants that remain infected then progresses into the Infectious (and Diseased) compartments after another latent or post-latent incubation period, $t_{i}^I, t_{i}^D$. These plants are then considered chronically infected are excluded from winter recovery for the rest of the simulation.


```{r}

hostRecovery <- function(time){
  (1 - exp(-0.045*time))^135.9
}

timestep <- 1:365
test <- hostRecovery(timestep)
qplot(x = timestep/7, y = test, geom = "path", xlab = "Week", ylab = "Probability of winter recovery")

```

\ 

#### Roguing and vector overwintering

Roguing and vector overwintering still need to be developed and incorporated into the model.

\ 

## Running the model

```{r, fig.cap = "Raster map of infection status of each plant", fig.align="center"}

#### Parameter values
## In this version, epsilon and beta are functions of other parameters, define non-varying parameters here
alpha <- 10 # Dispersal parameter
eta <- 0.5 # Inoculation rate (LAMBDA in Parry et al.)
kappa_e <- 0.6 # Proportion of external vectors infectious
a <- 0.6 # Acquisition rate
muv <- 0.1 # In-field loss rate of vectors (due to both death and emigration)
### Parameter values for vectorDensity() damped sine wave
A <- -10 # Initial amplitude
lambda_osc = 0.05 # Dampening
phi = 0.09 # Angular frequency
base = 0.1 # Baseline vector density

## Number of time steps; assume each time step is 1 week
Tmax <- 52
## Number of years
numYears <- 3

#### Plant spatial coordinates
## Total number of plants (numPlants) = nrow(Coo) = nrc*nrc
nrc <- 30
Coo <- makeCoordinates(nrc)

## Calculating distance to nearest field border for each plant
# borderCoo <- rep(c(1,nrc), each = 2)
# borderD <- apply(Coo, 1, function(x) min(sqrt((x - borderCoo)^2))+1)
## Distance from the left border whould just be the 1st coordinate
borderD <- Coo[,1]
## Add 1 to the distance; plants on the border are then 1 space from the border and all other plants are shifted accordingly
## Dispersal distances from border
## When d = 0, normalizedKernel = Inf
epsilonK <- chooseKernel(d = borderD, alpha = alpha, kernelFunc = "normalized exponential")


#### Setting up empty vectors and matrices to track things
## Assumes already created a grid of hosts
## Calculate total number of plants and vector of plant IDs
numPlants <- nrow(Coo)
IDs <- 1:numPlants
#### Set up vectors to track state of each plants
## Exposure times should be a matrix: rows = plants, columns = years
Exp_times <- matrix(Tmax, nrow = numPlants, ncol = numYears)
## Vectors of rho_epsilon(t), rho_beta(t), and beta(t) values
rho_etVec <- rho_btVec <- kappaVec <- betaVec <- rep(0, Tmax)
## Matrix of epsilon_i(t) values
epsilonMatrix <- matrix(NA, nrow = numPlants, ncol = Tmax)

#### Start year loop
for(y in 1:numYears){
  ####################################################################################################
  #### Stochastic transitions among host compartments
  ## (Re)set Infection times, Disease times (i.e., time of symptom onset)
  Inf_times <- Disease_times <- rep(Tmax, numPlants)
  ## (Re)set lambda values for each plant
  lambda <- rep(0, numPlants)
  ## (Re)set Sellke thresholds for infection
  Q <- rexp(numPlants, rate = 1)
  #### (re)set stochastic latent and cryptic periods
  ## Parry et al. (2014) assumed cryptic period (I -> D) is gamma distributed
  ## Here, we make a similar assumption for latent period (E -> I)
  ## Assume that each period has an average of 4 weeks
  ## a*s (mean) = 4; a:s (ratio) = 4:1; this ratio produces a humped distribution
  ## Parameterization assumes that time step = 1 week, will need to be adjusted if time step changes
  ## Durations of stays in Exposed compartment
  Et <- rgamma(numPlants, shape = 4, scale = 1)
  ## Durations of stays in Infectious compartment
  It <- rgamma(numPlants, shape = 4, scale = 1)
  #####################################################################################################
  #### Nested for loops of t, Susceptible plants, and Infected plants
  for(t in 1:Tmax){
    ## Each time point, update vectors of infecteds, susceptibles, and their respective indices
    Infected_ind <- which(Inf_times < Tmax)
    #Infected_times <- Inf_times[Infected_ind]
    Susceptible_ind <- which(Exp_times[,y] == Tmax)
    #Susceptible_times <- Inf_times[Susceptible_ind]
    ## Summaries of infections and susceptibles
    numInfections <- length(Infected_ind)
    numSusceptibles <- length(Susceptible_ind)
    ## Loop over susceptible plants
    for(iiSusceptible in Susceptible_ind){
      infTimeSusceptible <- Exp_times[iiSusceptible, y]
      CooSusceptible <- Coo[iiSusceptible,1:2]
      ## (Re)set dispersal kernel summation to 0
      m <- 0
      #### Loop over infected plants to get dispersal kernel from each Infected to a given Susceptible
      for(iInfected in Infected_ind) {
        infTimeInfected <- Inf_times[iInfected]
        if(infTimeSusceptible > infTimeInfected) { ## This avoids Target plant being same as Source
          ## Calculate d for each Infected plant
          d <- sum(sqrt((Coo[iInfected,1:2] - CooSusceptible)^2))
          ## Calculate the dispersal kernel based on each d and then sum over all kernels
          ## Using Neri et al. 2014 normalized exponential kernel
          ## When using simple exponential kernel, m becomes too large, lambdaii > 1, and rbinom fails with error
          m <- m + chooseKernel(distance = d, alpha = alpha, kernelFunc = "normalized exponential")
        }
        ## Evaluate if cryptic period is exceeded
        ## Infectious plants become Diseased when time step t exceeds Inf_times + It
        ## Only update Infectious plants that aren't Diseased yet
        if(t >= (infTimeInfected + It[iInfected]) & Disease_times[iInfected] == Tmax){
          if(Disease_times[iInfected] < t){
            warning("Old disease time being updated")
          }
          Disease_times[iInfected] <- runif(1, min = t-1, max = t)
        }
      } # End iInfected loop
      #### Epsilon calculations
      ## Seasonal vector density as damped sine wave
      rho_et <- vectorDensity(A = A, lambda_osc = lambda_osc, phi = phi, base = base, time = t) 
      ## Save rho_et
      rho_etVec[t] <- rho_et
      # rho_etSum <- sum(rho_etVec) ## Cumulative rho_et
      epsilonti <- eta*kappa_e*rho_et*epsilonK[iiSusceptible]
      ## Save epsilonti
      epsilonMatrix[iiSusceptible, t] <- epsilonti
      #### Beta calculations
      Infecteds_tm1 <- sum(Inf_times < t) # Number of infected plants at t-1
      atm1 <- holling4(a = 0.1, b = 72, c = -16, time = t-1) # Time-dependent acquisition rate
      kappa_t <- atm1*Infecteds_tm1/numPlants # Time-dependent in-field proportion vectors infectious
      kappaVec[t] <- kappa_t # Save kappa(t)
      ## In-field vector density
      ## if() statement for case of first time step
      if(t == 1){
        rho_bt_tm1 <- 0
        rho_et_tm1 <- 0
      } else {
        rho_bt_tm1 <- rho_btVec[t-1]
        rho_et_tm1 <- rho_etVec[t-1]
      }
      rho_btVec[t] <- (1 - muv)*(rho_bt_tm1 + rho_et_tm1) # Calculate in-field vector density
      beta_t <- eta*kappa_t*rho_btVec[t] # Calculate beta_t
      betaVec[t] <- beta_t # Save beta_t
      ## Cumulative force of infection for iiSusceptible plant
      lambda[iiSusceptible] <- lambda[iiSusceptible] + beta_t*m + epsilonti
      ## If cumulative lambda (i.e., infection pressure) exceeds Q (i.e., Sellke threshold) plant becomes infected,
      ## then receives an infection time in range [t - 1, t]
      if(lambda[iiSusceptible] >= Q[iiSusceptible]){
        Exp_times[iiSusceptible, y] <- runif(1, min = t-1, max = t)
      }
    } # End iiSusceptible loop
    ########################################################################################
    #### Moving from Exposed to Infectious compartments
    ## Select indices of plants that are in Exposed compartment (but not Infectious compartment yet)
    Exposed_ind <- which(Exp_times[,y] < Tmax & Inf_times == Tmax)
    ## Evaluate if latent period is exceeded
    ## Exposed plants become Infectious when time step t exceeds Exp_times + Et 
    for(iExposed in Exposed_ind){
      if(Inf_times[iExposed] < t){
        warning("Old infection time incorrectly being updated")
      }
      if(t >= Exp_times[iExposed, y] + Et[iExposed]){
        Inf_times[iExposed] <- runif(1, min = t-1, max = t)
      }
    } # End iExposed loop
  } # End t loop
  ###########################################################################################
  #### Winter processes
  #### Host recovery
  ## Base host recovery on Exposure time, rather than Infection time
  ## Get all plants that became Exposed; differs from Exposed_ind above
  expIndices <- which(Exp_times[,y] < Tmax)
  recoveryStatus <- probRecovery <- rep(0, length(numPlants))
  for(iExposed in expIndices){
    ## hostRecovery function is based on Feil et al. 2003 equation based on day of year; need to convert to week time step
    probRecovery[iExposed] <- hostRecovery(Exp_times[iExposed, y]*7)
    recoveryStatus[iExposed] <- rbinom(1, 1, probRecovery[iExposed])
    ## Make sure recoveryStatus is either 0 or 1
    if(recoveryStatus[iExposed] != 0 & recoveryStatus[iExposed] != 1){
      warning("Something is wrong with recoveryStatus")
    }
    ## For chronic infected plants, set Exp_times for next year to 0
    if(y < numYears & recoveryStatus[iExposed] == 0){
      Exp_times[iExposed, y+1] <- 0
    }
  } # End iExposed loop (for recovery)
  ## Check that recovery probability calculations are working correctly
  recoveryCheck <- cbind(Exp_times[expIndices], probRecovery[expIndices], recoveryStatus[expIndices])
} # End y loop

#### Produce raster with infection times as values
## Get raster dimensions from Coo
## Coo[,1] = rows or y coord
## Coo[,2] = columns or x coord
CooRows <- Coo[,1]
CooNrows <- length(unique(CooRows))
CooYmn <- min(CooRows)
CooYmx <- max(CooRows)
CooColumns <- Coo[,2]
CooNcols <- length(unique(CooColumns))
CooXmn <- min(CooColumns)
CooXmx <- max(CooColumns)

rasterTimes <- raster(nrows = CooNrows, ymn = CooYmn, ymx = CooYmx,
                      ncols = CooNcols, xmn = CooXmn, xmx = CooXmx,
                      vals = Disease_times)
plot(rasterTimes, main = "Week of year plant became diseased")


```

\ 

The simulation above was run for 1 year, with a time step of 1 week (i.e., 52 time steps), on a vineyard of 30 x 30 = 900 plants. In the figure above, green squares are healthy plants at the end of the year, with the other colors indicating the infection time of infected plants: lighter colors indicate earlier infection time. With the above parameter values, most infection occurs near the border apparently from immigrating vectors, rather than secondary spread. 
\ 
The figure below shows the number of cumulative infections over time

```{r, fig.cap = "Cumulative number of infections per month", fig.align = "center"}

## Vector of week number at the end of years, to produce vertical lines on ggplot
endOfYears <- data.frame(timeStep = Tmax*1:numYears)

## Vector of each time step (weeks)
weekVector <- data.frame(timeStep = 1:(Tmax*numYears))

## Round up infection times
finalInfectionTimes <- ceiling(Exp_times)

infectionTimesList <- vector("list", numYears)
for(i in 1:numYears){
  times.i <- finalInfectionTimes[,i]
  times.i <- times.i[times.i < Tmax]
  infCounts <- rle(sort(times.i))
  infectionTimesList[[i]] <- data.frame(timeStep = infCounts$values+(Tmax*(i-1)), 
                                        numInfections = infCounts$lengths,
                                        cumulInfections = cumsum(infCounts$lengths))
}
infectionTimes <- infectionTimesList %>% rbindlist() %>% as.data.frame() %>%
  ## Expand data set so that each time step is included explicitly
  right_join(., weekVector, by = "timeStep") %>%
  fill(cumulInfections) %>% # Replace NAs with previous non-NA value
  ## Replace NAs with 0
  ## Can replace NAs in cumulInfections with 0 now because these are just at the start of the time series
  replace_na(list(numInfections = 0, cumulInfections = 0)) 


ggplot(data = infectionTimes, aes(x = timeStep, y = cumulInfections)) +
  geom_path(lwd = 1) +
  ## Add vertical lines indicating the end of each growing season/year
  geom_vline(data = endOfYears, aes(xintercept = timeStep), linetype = 2) +
  scale_x_continuous(name = "Weeks") +
  scale_y_continuous(name = "Cumulative number of latent infections") +
  theme_bw(base_size=16) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(colour = "black"),
        panel.background = element_blank(),
        legend.position = "none") 

# hist(infectionTimes)
# infSummary <- Infections %>% floor() %>% table() %>% as.data.frame()
# names(infSummary) <- c("time", "numInfections")
# infSummary$cumulative <- cumsum(infSummary$numInfections)
# infSummary$time <- factor2numeric(infSummary$time)


```

\ 

#### Plot dynamics of in-field vector density and in-field vector infectivity

```{r, fig.align="center", fig.cap="In-field vector density over time"}

#### Beta values for each time step
plotRho_bt <- data.frame(rho_btVec,
                         week = 1:length(rho_btVec))

ggplot(data = plotRho_bt, aes(x = week, y = rho_btVec)) +
  geom_path(lwd = 1) +
  scale_x_continuous(name = "Week", limits = c(1,52), breaks = seq(1,52,by=5)) +
  scale_y_continuous(name = "In-field vector density") +
  theme_bw(base_size=16) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(colour = "black"),
        panel.background = element_blank(),
        legend.position = "none") 

```

\ 

#### Number of new infected plants scaled by in-field vector density

```{r, fig.align = "center", fig.cap = "Number of new infections scaled by vector density. Note, chronic infections were removed from figure"}

#### Add vector density to infection times data set
## rho_btVec is a vector of densities of length Tmax; as long as nrow(infectionTimes) is a multiple of Tmax, rho_btVec values will be recycled
infectionTimes$vectorDensity <- rho_btVec
infectionTimes$scaledInfections <- with(infectionTimes, numInfections/vectorDensity)

#### Remove chronic infections -- can be accomplished by removing the last week of each year
infectionTimes <- infectionTimes[!(infectionTimes$timeStep %in% endOfYears$timeStep),]

ggplot(data = infectionTimes, aes(x = timeStep, y = scaledInfections)) +
  geom_path(lwd = 1) +
  geom_vline(data = endOfYears, aes(xintercept = timeStep), linetype = 2) +
  scale_x_continuous(name = "Week") +
  scale_y_continuous(name = "Number of new latent infections per vector") +
  theme_bw(base_size=16) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(colour = "black"),
        panel.background = element_blank(),
        legend.position = "none") 

```


#### Dynamics of external and secondary infection rates

```{r, fig.align="center", fig.cap="Mean external infection rates over time"}

#### Average external infection rate (epsilon) over susceptible plants at each time step
plotEpsilon <- data.frame(meanEpsilon = colMeans(epsilonMatrix, na.rm = TRUE),
                          week = seq(1, 52, length.out = Tmax))

ggplot(data = plotEpsilon, aes(x = week, y = meanEpsilon)) +
  geom_path(lwd = 1) +
  scale_x_continuous(name = "Week", limits = c(1,52), breaks = seq(1,52,by=5)) +
  scale_y_continuous(name = "Mean epsilon") +
  theme_bw(base_size=16) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(colour = "black"),
        panel.background = element_blank(),
        legend.position = "none") 

```

\ 

```{r, fig.align="center", fig.cap="Secondary infection rate over time"}

#### Beta values for each time step
plotBeta <- data.frame(betaVec,
                       week = seq(1,52, length.out = length(betaVec)))

ggplot(data = plotBeta, aes(x = week, y = betaVec)) +
  geom_path(lwd = 1) +
  scale_x_continuous(name = "Week", limits = c(1,52), breaks = seq(1,52,by=5)) +
  scale_y_continuous(name = "Beta") +
  theme_bw(base_size=16) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(colour = "black"),
        panel.background = element_blank(),
        legend.position = "none") 

```

\ 

```{r, fig.align="center", fig.cap="In-field vector infectivity over time"}

#### Beta values for each time step
plotkappa_t <- data.frame(kappaVec,
                          week = seq(1,52, length.out = length(kappaVec)))

ggplot(data = plotRho_bt, aes(x = week, y = kappaVec)) +
  geom_path(lwd = 1) +
  scale_x_continuous(name = "Week", limits = c(1,52), breaks = seq(1,52,by=5)) +
  scale_y_continuous(name = "Proportion of in-field vectors infectious") +
  theme_bw(base_size=16) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(colour = "black"),
        panel.background = element_blank(),
        legend.position = "none") 

```


Next steps:  
1. The parameter values are totally made up at this point. I/we would need to think about the time steps and what appropriate parameter values should look like.  
2. I'm sure I could improve upon the damped sine wave for vector density. This is a "first pass". Suggestions would be great.  
3. Incorporating winter recovery and roguing as described in the Powerpoint slides I sent awhile ago.